---
title: "Analysis of US Household Incomes"
author: "Shreyas Srinath"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
---

<h3>1) Introduction </h3>
<br />


The United States has routinely collected information regarding household
incomes across multiple states, districts, counties and cities. This data
can then be further analyzed to provide a snapshot of the state of modern
American households.

By analyzing data found on Kaggle from the year 2017, I have attempted to 
break down a rich dataset into its constituent parts to uncover patterns,
insights and interesting facts about the US as it pertains to annual incomes.

<br />
<h3>2) Data </h3>
<br />

**Citation:** Golden Oak Research Group, LLC. “U.S. Income Database Kaggle”. <br />
**Publication:** 5, August 2017. <br />
**Accessed:** 13, December 2022. <br />

The data is obtained from a report of *Golden Oak Research Group*, which has
partitioned a larger dataset to be used by city planners and real estate 
developers. 

The dataset comprises 30,000+ rows of household income data with geographical,
regional and statistical indicators.

The following attributes are clearly defined:

1) <var>id</var> : A random row identifier (character) <br />
2) <var>State_Code</var> : State identifier (character) <br />
3) <var>State_Name</var> : State name (character) <br />
4) <var>State_ab</var> : State Code (character) <br />
5) <var>County</var> : County Name (character) <br />
6) <var>City</var> : City Name (character) <br />
7) <var>Place</var> : Place Name (character) <br />
8) <var>Type</var> : Place Type (character). Can be: <br />
                                -> Borough <br />
                                -> CDP <br />
                                -> City <br />
                                -> Community <br />
                                -> County <br />
                                -> Municipality <br />
                                -> Town <br />
                                -> Track <br />
                                -> Urban <br />
                                -> Village <br />
9) <var>Primary</var> : Primary Type (character) <br />
10) <var>Zip_Code</var> : Zip Code (character) <br />
11) <var>Area_Code</var> : Area Code (double) <br />
12) <var>ALand</var> : Land Area (double) <br />
13) <var>AWater</var> : Water Area (double) <br />
14) <var>Lat</var> : Latitude Coordinate (double) <br />
15) <var>Lon</var> : Longitude Coordinate (double) <br />
16) <var>Mean</var> : Mean Household Income in $ (double) <br />
17) <var>Median</var> : Median Household Income in $ (double) <br />
18) <var>Stddev</var> : Standard Deviation in $ (double) <br />
19) <var>sum_w</var> : Unidentified metric (double) <br />

<br />
<h3>3) Importing Libraries </h3> <br />
I have employed the <var>tidyverse</var>, <var>ggplot2</var>, <var>dplyr</var>,
<var>forcats</var> and <var>mapview</var> libraries for the analysis. <br />
<br />

```{r setup, warning = FALSE}

#######################
# Importing Libraries #
#######################

library(tidyverse)
library(sf)
library(ggplot2)
library(mapview) # Used to plot clusters geographically
library(dplyr)
library(fastDummies) # Used to generate dummy variables
library(forcats)

```

<h3>4) Importing Dataset </h3> <br />
I have used the <var>read_csv</var> method to import the data as it is stored in 
the *Comma Separated Values* format. <br />
<br />

```{r, warning = FALSE}

#####################
# Importing Dataset #
#####################

dataset <- as.data.frame(read_csv("kaggle_income.csv"))
# Display first six rows
head(dataset)

```

<br />
<h3>5) Tidying Dataset </h3> <br />
I have identified a single observational row with a missing value. Since the data
that is missing is qualitative, I decided to omit that row from further 
analysis. <br /> 
<br />

```{r, warning = FALSE}
##########################
# Handling Missing Values#
##########################

colSums(is.na(dataset))
# Identify the indices for the missing value
which(is.na(dataset), arr.ind = TRUE)
# Remove that entry from the dataset
dataset <- dataset[-27548,]
# Check for missing values
colSums(is.na(dataset))
```

<br />
<h3>6) Preliminary Analysis </h3> <br />
This is done for understanding the dimensionality the levels for
the qualitative variables and correlations for the quantiative variables. <br />
<br />

```{r, warning = FALSE}

############################
# Understanding the Dataset#
############################

dim(dataset)
# Dataset has 32525 rows and 19 variables

# Check Levels and Count for qualitative variables

# - State_Name
sort(unique(dataset$State_Name))
length(unique(dataset$State_Name))

# - Type
sort(unique(dataset$Type))
length(unique(dataset$Type))

# - Primary
sort(unique(dataset$Primary))
length(unique(dataset$Primary))

########################
# Checking Correlations#
########################

cor(dataset[c(12:17)])
# Slight positive correlation between Latitude and Mean Income

```

<br />
<h3>7) Region-wise Analysis </h3> <br />

<br />
In order to summarise statistics for a given region, I have employed forcats
to lump the states according to their respective geographical regions
within the US. 

Here is a brief overview of state-wise distribution of incomes: <br /> 
<br />

```{r, warning = FALSE}
#############################
# Grouping States by Region #
#############################

dataset%>%
  mutate(region = fct_collapse(State_Name,
                               NewEngland = c("Connecticut", "Maine", 
                                              "Massachusetts", "New Hampshire", 
                                              "Rhode Island", "Vermont"),
                               MidAtlantic = c("Delaware", "Maryland", 
                                               "New Jersey", "New York", 
                                               "Pennsylvania", "Virginia"),
                               South = c("Alabama", "Arkansas", "Florida", 
                                         "Georgia", "Kentucky", "Louisiana", 
                                         "Mississippi", "Missouri", 
                                         "North Carolina", "South Carolina", 
                                         "Tennessee",  
                                         "West Virginia"),
                               Midwest = c("Illinois", "Indiana", "Iowa", 
                                           "Kansas", "Michigan", "Minnesota", 
                                           "Nebraska", "North Dakota", "Ohio", 
                                           "South Dakota", "Wisconsin"),
                               SouthWest = c("Arizona", "New Mexico", 
                                             "Oklahoma", "Texas"),
                               West = c("Alaska", "California", "Colorado", 
                                        "Hawaii", "Idaho", "Montana", "Nevada", 
                                        "Oregon", "Utah", "Washington", 
                                        "Wyoming"))) -> state_data
unique(state_data$region)


##########################
# Statewise Total Incomes#
##########################

NewEngland<-state_data%>%
  select(State_ab,County,City,region,Median)%>%
  group_by(region == "NewEngland",State_ab)

NewEngland%>%
  group_by(State_ab)%>%
  summarise(total=sum(Median)) -> NewEnglandSummary


ggplot(NewEnglandSummary, aes(x =reorder(State_ab,desc(total)), 
                       y=total,
                       fill=total)) + 
  geom_bar(stat="identity", show.legend = FALSE)+
  scale_y_continuous(n.breaks = 16, labels=scales::comma) +
  scale_fill_gradient(low="blue", high="red") +
  xlab("State") + ylab("Statewise Total Income in $") +
  ggtitle("Total Incomes by State")


```

<br />
<h4>7.1) Grouping US States by Avg Median Income </h4> <br />
Summarising US Regions by Average Median Income helps us understand which 
portions of the US are earning higher on average.

The results are discussed below: <br /> 
<br />


```{r, warning = FALSE}
#######################################
# Grouping US Regions by Median Income#
#######################################

state_data%>%
  group_by(region)%>%
  summarise(AvgMedianInc = mean(Median))%>%
  ggplot(mapping = aes(x=AvgMedianInc, y=fct_reorder(region, AvgMedianInc))) +
  geom_bar(stat = "identity", width=0.7, show.legend = FALSE, fill = "#85bb65") +
  coord_flip() +
  geom_text(aes(label = round(AvgMedianInc, digits = 2)), vjust = 2,
            size = 3, color = "white") +
  xlab("Region") +
  ylab("Average Median Income in $") +
  ggtitle("US Regions ordered by Avg. Median Income in $") 

```

<br />
<h4>7.2) Discussion for New England States</h4> <br />
We dive deeper to New England States to understand which states tend to
outperform others on average.

The results are discussed below: <br /> 
<br />

```{r, warning = FALSE}

###############
# NEW ENGLAND #
###############

# Filter observational rows for New England States
state_data%>%
  filter(region == "NewEngland")%>%
  group_by(State_Name, State_ab)%>%
  summarise(StateAvgInc = mean(Median))%>%
  arrange(desc(StateAvgInc)) -> NewEnglandData

# Show Data for New England States
head(NewEnglandData)

# Plot New England States based on Avg Median Income
ggplot(NewEnglandData, mapping = aes(x = reorder(State_Name, StateAvgInc), 
                                     y = StateAvgInc,
                                     fill = State_ab)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_y_continuous(n.breaks = 16, labels=scales::comma) +
  geom_text(aes(label = round(StateAvgInc, digits = 2)), vjust = 2,
            size = 4, color = "white") +
  xlab("State") +
  ylab("Average Earnings of State in $") +
  ggtitle("New England States ordered by Avg Median Income in $") 

```
<br />
**Connecticut** and **Massachusetts** dominate the households with higher average
incomes since these states have the highest concentration of capital and
industries whereas the likes of Rhode Island, Vermont, New Hampshire and Maine
rely on income from tourism which is seasonal.
<br /> 

<br />
<h4>7.3) Discussion for Mid Atlantic States</h4> <br />
Checking which Mid Atlantic States tend to outperform other states on average.

The results are discussed below: <br /> 
<br />

```{r, warning = FALSE}
################
# MID ATLANTIC #
################

# Filter observational rows for only Mid Atlantic States 
state_data%>%
  filter(region == "MidAtlantic")%>%
  group_by(State_Name, State_ab)%>%
  summarise(StateAvgInc = mean(Median))%>%
  arrange(desc(StateAvgInc)) -> MidAtlanticData

# Display first six rows of filtered data
head(MidAtlanticData)

# Plot states in Mid Atlantic region by Avg Median Income
ggplot(MidAtlanticData, mapping = aes(x = reorder(State_Name, StateAvgInc), 
                                     y = StateAvgInc,
                                     fill = State_ab)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_y_continuous(n.breaks = 16, labels=scales::comma) +
  geom_text(aes(label = round(StateAvgInc, digits = 2)), vjust = 2,
            size = 4, color = "white") +
  xlab("State") +
  ylab("Average Earnings of State in $") +
  ggtitle("Mid Atlantic States ordered by Avg Median Income in $")

```

<br />
**New Jersey** and **Maryland** dominate the households with higher average
incomes but the distribution of household income remains relatively uniform.
<br /> 


<br />
<h4>7.3) Discussion for Southern States</h4> <br />
Checking which Mid Atlantic States tend to outperform other states on average.

The results are discussed below: <br /> 
<br />

```{r, warning = FALSE}
################
# SOUTH REGION #
################

# Filter observational rows for Southern States
state_data%>%
  filter(region == "South")%>%
  group_by(State_Name, State_ab)%>%
  summarise(StateAvgInc = mean(Median))%>%
  arrange(desc(StateAvgInc)) -> SouthData

# Display first 6 rows
head(SouthData)

# Plot southern states by Avg Median Income
ggplot(SouthData, mapping = aes(x = reorder(State_ab, StateAvgInc), 
                                y = StateAvgInc,
                                fill = State_ab)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_y_continuous(n.breaks = 16, labels=scales::comma) +
  geom_text(aes(label = round(StateAvgInc, digits = 2)), vjust = 2,
            size = 2, color = "white") +
  xlab("State") +
  ylab("Average Earnings of State in $") +
  ggtitle("Southern States ordered by Avg Median Income in $") 

```

<br />
**Florida** and **Georgia** dominate the households with higher average
incomes but the distribution of household income remains relatively uniform.
<br /> 

<br />
<h3> 8) State-wise Analysis</h3> <br />
Plotting states on the basis of their Average Median Incomes reveals a
clear dichotomy between Lower Income, Middle Income and Higher Income states
as shown. <br /> 
<br />

```{r, warning = FALSE}
############################################
# Grouping States by Average Median Income #
############################################

dataset%>%
  group_by(State_Name)%>%
  summarise(AvgMedians = mean(Median))%>%
  ggplot(mapping=aes(x=AvgMedians, y= fct_reorder(State_Name, AvgMedians))) + 
  geom_point() +
  xlab("Average of State's Median Income ($)") +
  ylab("State") +
  ggtitle("States Ordered by Median Income (in Avg $ per State)") +
  theme(text = element_text(size = 10), element_line(size = 1)) 
```

<br />
<h4> 8.1) Analysis for New York State</h4> <br />
BY further looking at NY Counties, we can begin to analyze which counties
perform better and plot them geographically to realize where high earning
clusters lie.<br /> 
<br />

```{r, warning = FALSE}

###############################
# Analysis for New York State #
###############################

# Subset the dataset using dplyr to obtain rows for New York
dataset%>%
  filter(State_ab == "NY") -> datasetNY
# Display first six rows
head(datasetNY)

# Summarise the NY Data based on Latitudes, Longitudes and Average Income
# for the top 10 earning counties
datasetNY%>%
  group_by(County)%>%
  summarise(AvgLat = mean(Lat),
            AvgLon = mean(Lon),
            AvgInc = median(Median))%>%
  arrange(desc(AvgInc))%>%
  slice(1:10) -> sumdatasetNY


# Display first six rows of NY Data
head(sumdatasetNY)

# Plot the Avg. Median Income for each county in NY in descending order
sumdatasetNY%>%
  ggplot(mapping=aes(x=AvgInc, y= fct_reorder(County, AvgInc), fill= AvgInc)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_fill_gradient2(high = 'blue', mid = 'snow', low = 'red',
                       midpoint = median(sumdatasetNY$AvgInc)) +
  xlab("Average of NY's Median Income ($)") +
  ylab("NY Counties") +
  ggtitle("NY Counties Ordered by Median Income (in Avg $)") 
  
# Plot the counties on a map with markers to indicate highest earning counties
mapview(sumdatasetNY, xcol = "AvgLon", ycol = "AvgLat", crs = 4269, grid=FALSE,
        cex = "AvgInc", burst=TRUE, legend = TRUE)


# Plot earnings by Type of location
datasetNY%>%
  group_by(Type)%>%
  summarise(AvgEarnings = mean(Median))%>%
  arrange(desc(AvgEarnings))%>%
  ggplot(mapping=aes(x=AvgEarnings, y=fct_reorder(Type, AvgEarnings),
                     fill = Type)) + 
  geom_bar(stat = "identity",width=0.5, show.legend = FALSE) + coord_flip() +
  geom_text(aes(label = round(AvgEarnings, digits = 2)), vjust = 2,
            size = 4, color = "white") +
  xlab("Type of County") +
  ylab("Average Earnings of County in $") +
  ggtitle("NY County Types ordered by Avg. Earnings in $") 

```

<br />
<h4> 8.2) Analysis for Maine State</h4> <br />
BY further looking at ME Counties, we can begin to analyze which counties
perform better and plot them geographically to realize where high earning
clusters lie.<br /> 
<br />

```{r, warning = FALSE}

############################
# Analysis for Maine State #
############################

# Subset the dataset using dplyr to obtain rows for Maine
dataset%>%
  filter(State_ab == "ME") -> datasetME
# Display first six rows
head(datasetME)

# Summarise the NY Data based on Latitudes, Longitudes and Average Income
# for the top earning counties
datasetME%>%
  group_by(County)%>%
  summarise(AvgLat = mean(Lat),
            AvgLon = mean(Lon),
            AvgInc = median(Median))%>%
  arrange(desc(AvgInc))-> sumdatasetME

# Display first six rows of ME Data
head(sumdatasetME)

# Plot the Avg. Median Income for each county in ME in descending order
sumdatasetME%>%
  ggplot(mapping=aes(x=AvgInc, y= fct_reorder(County, AvgInc), fill= AvgInc)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_fill_gradient2(high = 'blue', mid = 'snow', low = 'yellow',
                       midpoint = median(sumdatasetME$AvgInc)) +
  xlab("Average of ME's Median Income ($)") +
  ylab("ME Counties") +
  ggtitle("ME Counties Ordered by Median Income (in Avg $)") 

# Plot the counties on a map with markers to indicate highest earning counties
mapview(sumdatasetME, xcol = "AvgLon", ycol = "AvgLat", crs = 4269, grid=FALSE,
        cex = "AvgInc", burst=TRUE, legend = TRUE)


# Plot earnings by Type of location
datasetME%>%
  group_by(Type)%>%
  summarise(AvgEarnings = mean(Median))%>%
  arrange(desc(AvgEarnings))%>%
  ggplot(mapping=aes(x=AvgEarnings, y=fct_reorder(Type, AvgEarnings),
                     fill = Type)) + 
  geom_bar(stat = "identity",width=0.5, show.legend = FALSE) + coord_flip() +
  geom_text(aes(label = round(AvgEarnings, digits = 2)), vjust = 2,
            size = 4, color = "white") +
  xlab("Type of County") +
  ylab("Average Earnings of County in $") +
  ggtitle("ME County Types ordered by Avg. Earnings in $") 

```

<br />
<h4> 8.3) Analysis for Texas State</h4> <br />
BY further looking at TX Counties, we can begin to analyze which counties
perform better and plot them geographically to realize where high earning
clusters lie.<br /> 
<br />

```{r, warning = FALSE}

############################
# Analysis for Texas State #
############################

# Subset the dataset using dplyr to obtain rows for Maine
dataset%>%
  filter(State_ab == "TX") -> datasetTX
# Display first six rows
head(datasetTX)

# Summarise the TX Data based on Latitudes, Longitudes and Average Income
# for the top earning counties
datasetTX%>%
  group_by(County)%>%
  summarise(AvgLat = mean(Lat),
            AvgLon = mean(Lon),
            AvgInc = median(Median))%>%
  slice(1:15)%>%
  arrange(desc(AvgInc))-> sumdatasetTX

# Display first six rows of TX Data
head(sumdatasetTX)

# Plot the Avg. Median Income for each county in TX in descending order
sumdatasetTX%>%
  ggplot(mapping=aes(x=AvgInc, y= fct_reorder(County, AvgInc), fill= AvgInc)) + 
  geom_bar(stat="identity", show.legend = FALSE) +
  scale_fill_gradient2(high = 'blue', mid = 'red', low = 'white',
                       midpoint = median(sumdatasetTX$AvgInc)) +
  xlab("Average of TX's Median Income ($)") +
  ylab("TX Counties") +
  ggtitle("TX Counties Ordered by Median Income (in Avg $)") 

# Plot the counties on a map with markers to indicate highest earning counties
mapview(sumdatasetTX, xcol = "AvgLon", ycol = "AvgLat", crs = 4269, grid=FALSE,
        cex = "AvgInc", burst=TRUE, legend = TRUE)


# Plot earnings by Type of location
datasetTX%>%
  group_by(Type)%>%
  summarise(AvgEarnings = mean(Median))%>%
  arrange(desc(AvgEarnings))%>%
  ggplot(mapping=aes(x=AvgEarnings, y=fct_reorder(Type, AvgEarnings),
                     fill = Type)) + 
  geom_bar(stat = "identity",width=0.5, show.legend = FALSE) + coord_flip() +
  geom_text(aes(label = round(AvgEarnings, digits = 2)), vjust = 2,
            size = 3, color = "white") +
  xlab("Type of County") +
  ylab("Average Earnings of County in $") +
  ggtitle("TX County Types ordered by Avg. Earnings in $") 

```
<br />
<h3>9) Regression Analysis </h3> <br />

Regression Analysis is used to fit the data for a given dependent variable in
terms of the other independent variables to estimate values of the dependent
variable given the other parameters.

We have employed the <var>fastDummies</var> library to initially convert
qualitative information into a set of dummy variables (0/1) to indicate the 
absence or presence of a quality respectively.

We have then performed Regression analysis using Median as the dependent 
variable. 

```{r}
# Reading updated dataset
regdata<-read.csv("regdata.csv")
# Replacing qualitative variables with dummy data
dumdata<-dummy_cols(regdata,select_columns = c("County","State_Name","Type"))
# Generating Regression Model with Median income as dependent variable
MLRmodel <- lm(Median ~ State_Name + Type + ALand, data=dumdata)
# Summary Results for Regression
summary(MLRmodel)

```

<br />
Regression analysis reveals that a State's Type and Area of Land play
a minimal role in determining Median Income.

We would require household incomes, industry types and migration details 
to conduct a more successful analysis using more powerful algorithms.
<br />

<br />
<h3>10) Conclusion </h3> <br />

By analyzing US household incomes, we have identified the top performing regions
as the *New England* and *Mid Atlantic* regions.

New England States are dominated by *Massachusetts* and *Connecticut* having
far more significant earnings than the other states.

Mid Atlantic and Southern States show a more even distribution of household 
median income.

Conducting a more in-depth analysis for individual states (NY, ME and TX),
we could identify clusters where there are high earning counties which can be
beneficial for city planners, real estate developers or out-of-state migrants
hoping to move into a particular region.

While Regression Analysis could not reveal further insight, we hope that in the
future, with details for households and industries that we can employ more
powerful analysis techniques.

</br>
</body>
</html>